Decoupled Fallback: A Portable Single-Pass GPU Scan

### 2025-06-19 paper.tex
Reorder figures, remove progress classification section.

\setcopyright{acmlicensed}
\copyrightyear{2025}
\acmYear{2025}
\acmDOI{XXXXXXX.XXXXXXX}
\acmConference[SPAA '25]{37th ACM Symposium on Parallelism in Algorithms and Architectures}{July 28--August 1,
  2025}{Portland, OR, USA}
\acmISBN{978-1-4503-XXXX-X/18/06}

@@

\begin{figure*}
  \centering
  \makebox[\textwidth][c]{
    \begin{subfigure}[c]{0.48\textwidth}
      \centering
      \includegraphics[height=7.5cm,keepaspectratio]{graphics/ScanThenPropagate.pdf}
      \caption{\emph{Scan-then-Propagate.}}
      \label{fig:scan-then-propagate}
    \end{subfigure}
    \hfill
    \begin{subfigure}[c]{0.48\textwidth}
      \centering
      \includegraphics[height=6cm,keepaspectratio]{graphics/ReduceThenScan.pdf}
      \caption{\emph{Raking Reduce-then-Scan.}}
      \label{fig:reduce-then-scan}
    \end{subfigure}%
  }\\[1em]

@@

@@

Moreover, the fused nature of \emph{Chained Scan} enables in-situ compaction: algorithms requiring compaction can merge that work directly into the scan kernel, avoiding redundant writes or secondary compaction passes. This fusion benefits applications ranging from data deduplication to compression. However, \emph{Chained Scan} has a critical limitation: its reliance on a forward-progress guarantee. While NVIDIA GPUs provide this guarantee, data collected by Sorensen et al.~\cite{sorensen2021} shows that ARM and Apple GPUs do not. Without forward-progress guarantees, \emph{Chained Scan} is not merely inefficient—it risks indefinite starvation, causing the application to freeze and rendering the kernel unable to terminate, requiring external intervention to recover.

\subsection{Progress Guarantees and Their Impact on \emph{Chained Scan}}
 Prior to Sorensen et al.~\cite{sorensen2018,sorensen2021}, developers largely relied on untested, folklore-like assumptions about GPU progress guarantees. Instead, Sorensen et al.\ systematically proposed and tested three progress models:\footnote{Note, a ``workgroup'' in Sorensen et al.'s definition is not necessarily in-flight and occupant on a multiprocessor.}
\begin{itemize}
  \item \textbf{Occupancy-Bound Execution (OBE)}: Workgroups that have made at least one execution step.
  \item \textbf{Heterogeneous System Architecture~\cite[Section~2.11]{HSASysArch12} (HSA)}: The workgroup with the lowest ID that has not yet terminated.
  \item \textbf{Linear OBE (LOBE)}: Workgroups that have made at least one step, and any workgroup with a lower ID than a workgroup that has made at least one step.
\end{itemize}
Their 2021 findings~\cite{sorensen2021} demonstrated that Apple and ARM GPUs do not support LOBE, contradicting their prior hypothesis~\cite{sorensen2018}.

Sorensen et al.\ classify \emph{Chained Scan} as a \emph{one-way producer-consumer} (PC) HSA algorithm: a workgroup only consumes from lower-ID counterparts, ensuring that execution proceeds in a strictly ordered fashion. HSA guarantees termination, as the lowest ID workgroup is always given forward progress. If subgroups from the lowest ID workgroup are not guaranteed execution, then dependent workgroups may spin indefinitely, waiting on an update that never arrives.

However, the HSA classification relies on the assumption that \emph{Chained Scan} follows scheduler-assigned virtualized IDs rather than dynamically acquired atomic-bumped IDs, as used by both \emph{StreamScan} and \emph{Decoupled Lookback}. Since virtual IDs do not necessarily match atomic IDs, HSA does not guarantee termination---the lowest virtual ID may not be scheduled first, potentially leading to starvation.

LOBE adds an additional condition on top of OBE\@: lower-ID workgroups must be scheduled before higher-ID ones. However, this condition is unnecessary for \emph{Chained Scan} for two reasons: (1) We already guarantee serial ordering of work items through atomic bumping, making OBE sufficient. (2) \emph{Chained Scan} only requires forward progress after a workgroup has atomically acquired its tile index. By definition, a workgroup that has acquired its index must be occupant on a physical processor (i.e., it has made at least one execution step); thus, the OBE guarantee is sufficient for forward progress.

Thus, by dynamically acquiring IDs through atomics, \emph{Chained Scan} shifts its progress model requirement from HSA to OBE. While LOBE encompasses both HSA and OBE, its stricter ordering guarantees are unnecessary for correctness. Ultimately, regardless of a device's specific progress model, any architecture lacking forward-progress guarantees will fail to execute \emph{Chained Scan}. In short:

\begin{table}[h!]
  \small
  \centering
  \begin{tabular}{lcccc}
    \toprule
    Algorithm                             & LOBE & OBE & HSA & None \\
    \midrule
    \emph{Chained Scan} with Workgroup ID & Yes  & No  & Yes & No   \\
    \emph{Chained Scan} with Atomic ID    & Yes  & Yes & No  & No   \\
    Ours                                  & Yes  & Yes & Yes & Yes  \\
    \bottomrule
  \end{tabular}
  \caption{Comparison of \emph{Chained Scan} Variants and Our Implementation Across Progress Models.\label{tab:progress_models}}
\end{table}

\subsection{Why is Reliance on Forward-Progress Guarantees a Portability Problem?}
A \emph{forward-progress guarantee} ensures that all occupant subgroups---of possibly different workgroups---are eventually given execution and make progress toward termination, preventing indefinite starvation. In the most comprehensive 
The absence of FPG in key hardware segments poses a critical challenge for developers. Our testing (Figure~\ref{fig:dist}) shows that, at best, running \emph{Chained Scan} without FPG results in performance worse than \emph{Reduce-then-Scan}, and frequently triggers \emph{Timeout Device Recovery}\label{sec:tdr}\footnote{Operating systems handle unresponsive programs differently. On Windows, the Timeout Detection and Recovery (TDR) system resets the GPU after approximately 2 seconds. Other operating systems---such as macOS and Linux---employ their own timeout mechanisms with varying behaviors. For consistency, we adopt Windows’s 2-second limit as a reasonable expectation for maximum execution time.} (TDR), crashing the program.
\begin{figure*}
  \centering
  \begin{subfigure}[b]{0.45\linewidth}
    \includegraphics[width=\linewidth]{graphics/decoupled_lookback.pdf}
    \caption{\emph{Decoupled Lookback.}}
    \label{fig:decoupled}
  \end{subfigure}
  \hfill
  \begin{subfigure}[b]{0.45\linewidth}
    \includegraphics[width=\linewidth]{graphics/ours.pdf}
    \caption{\emph{Decoupled Fallback (ours).}}
    \label{fig:ours}
  \end{subfigure}
  \caption{Histogram plots, 500 runs of Inclusive Prefix Sum, $\mathbf{2^{25}}$ 32-bit elements, on an FPG-lacking device, Apple M1 Max. Bar heights ($y$ axis) indicate the number of runs that completed in the execution time indicated on the $x$ axis. The Reduce-then-Scan threshold is the expected execution time of RTS at the same input size; runs to the left of this threshold indicate a performance uplift and vice versa. The TDR threshold is Windows's threshold for resetting the GPU\@; as noted in Section~\ref{sec:tdr}, MacOS handles this differently.}
  \label{fig:dist}
\end{figure*}
Exacerbating this issue, no major graphics API provides a mechanism to query FPG support. This lack of transparency forces developers to design algorithms without knowing whether critical guarantees are available on the target hardware. In shader environments, where cross-platform portability is essential, this uncertainty compels a fallback to older, less efficient strategies like \emph{Reduce-then-Scan}, sacrificing performance for broader compatibility.

@@

@@

\begin{figure*}
  \centering
  \includegraphics[width=.95\textwidth]{graphics/DecoupledFallback.pdf}
  \caption{\emph{Decoupled Fallback}. For brevity, the chained propagation phase is condensed, see \ref{fig:decoupled-lookback} for the full behavior.}
  \label{fig:decoupled-fallback}
\end{figure*}

@@

We now detail the \emph{Decoupled Fallback} procedure, presenting the algorithm from the perspective of a single workgroup. The details of tile state encoding and atomic updates are discussed in the following section and may vary depending on the data type, so for simplicity, we describe the operation as being performed by the \emph{first thread} of the workgroup:
\begin{figure}
  \centering
  \includegraphics[width=\columnwidth]{graphics/FlowChart.pdf}
  \caption{Flow Chart Representation of \emph{Decoupled Fallback}.\label{fig:decoupled_fallback}}
\end{figure}
\begin{enumerate}
  \item[(0)] \textbf{Posting Local Reduction:}
        \begin{itemize}
          \item Each workgroup posts its local reduction---the result of the workgroup-wide scan over partials---to global memory and updates its tile state to \emph{Ready}.
          \item If it is the first work tile, its local reduction \emph{is} its inclusive reduction. In this case, the workgroup updates its tile state directly to \emph{Inclusive} and bypasses the \emph{Chained Propagation} phase entirely.
        \end{itemize}

@@

>>>>

\acmYear{2025}\copyrightyear{2025}
\setcopyright{rightsretained}
\acmConference[SPAA '25]{37th ACM Symposium on Parallelism in Algorithms and Architectures}{July 28--August 1, 2025}{Portland, OR, USA}
\acmBooktitle{37th ACM Symposium on Parallelism in Algorithms and Architectures (SPAA '25), July 28--August 1, 2025, Portland, OR, USA}
\acmDOI{10.1145/3694906.3743326}
\acmISBN{979-8-4007-1258-6/25/07}

@@

\begin{figure*}[!t]
  \centering
  \makebox[\textwidth][c]{
    \begin{subfigure}[c]{0.48\textwidth}
      \centering
      \includegraphics[height=7.5cm,keepaspectratio]{graphics/ScanThenPropagate.pdf}
      \caption{\emph{Scan-then-Propagate.}}
      \label{fig:scan-then-propagate}
    \end{subfigure}
    \hfill
    \begin{subfigure}[c]{0.48\textwidth}
      \centering
      \includegraphics[height=6cm,keepaspectratio]{graphics/ReduceThenScan.pdf}
      \caption{\emph{Raking Reduce-then-Scan.}}
      \label{fig:reduce-then-scan}
    \end{subfigure}%
  }\\[1em]

@@

\begin{figure*}
  \centering
  \begin{subfigure}[b]{0.45\linewidth}
    \includegraphics[width=\linewidth]{graphics/decoupled_lookback.pdf}
    \caption{\emph{Decoupled Lookback.}}
    \label{fig:decoupled}
  \end{subfigure}
  \hfill
  \begin{subfigure}[b]{0.45\linewidth}
    \includegraphics[width=\linewidth]{graphics/ours.pdf}
    \caption{\emph{Decoupled Fallback (ours).}}
    \label{fig:ours}
  \end{subfigure}
  \caption{Histogram plots, 500 runs of Inclusive Prefix Sum, $\mathbf{2^{25}}$ 32-bit elements, on an FPG-lacking device, Apple M1 Max. Bar heights ($y$ axis) indicate the number of runs that completed in the execution time indicated on the $x$ axis. The Reduce-then-Scan threshold is the expected execution time of RTS at the same input size; runs to the left of this threshold indicate a performance uplift and vice versa. The TDR threshold is Windows's threshold for resetting the GPU\@; as noted in Section~\ref{sec:tdr}, MacOS handles this differently.}
  \label{fig:dist}
\end{figure*}

@@

Moreover, the fused nature of \emph{Chained Scan} enables in-situ compaction: algorithms requiring compaction can merge that work directly into the scan kernel, avoiding redundant writes or secondary compaction passes. This fusion benefits applications ranging from data deduplication to compression. However, \emph{Chained Scan} has a critical limitation: its reliance on a forward-progress guarantee. While NVIDIA GPUs provide this guarantee, data collected by Sorensen et al.~\cite{sorensen2021} shows that ARM and Apple GPUs do not. Without forward-progress guarantees, \emph{Chained Scan} is not merely inefficient---it risks indefinite starvation, causing the application to freeze and rendering the kernel unable to terminate, requiring external intervention to recover.

\subsection{Why is Reliance on Forward-Progress Guarantees a Portability Problem?}
\emph{Chained Scans} are \emph{one-way producer-consumer} algorithms; each workgroup consumes reductions only from lower-ID counterparts, thereby enforcing the ordering required to elide barriers. For such algorithms, guaranteeing execution of the lowest-ID workgroup is sufficient to ensure termination. If subgroups from the lowest-ID workgroup are not guaranteed execution, dependent workgroups may spin indefinitely, waiting for updates that never arrive. A \emph{forward-progress guarantee} ensures that all occupant subgroups---potentially from different workgroups---are eventually granted execution, thus preventing indefinite starvation.

The absence of FPG in key hardware segments poses a critical challenge for developers. Our testing (Figure~\ref{fig:dist}) shows that, at best, running \emph{Chained Scan} without FPG results in performance worse than \emph{Reduce-then-Scan}, and frequently triggers \emph{Timeout Device Recovery}\label{sec:tdr}\footnote{Operating systems handle unresponsive programs differently. On Windows, the Timeout Detection and Recovery (TDR) system resets the GPU after approximately 2 seconds. Other operating systems—such as macOS and Linux—employ their own timeout mechanisms with varying behaviors. For consistency, we adopt Windows’s 2-second limit as a reasonable expectation for maximum execution time.} (TDR), crashing the program. Exacerbating this issue, no major graphics API provides a mechanism to query FPG support. This lack of transparency forces developers to design algorithms without knowing whether critical guarantees are available on the target hardware. In shader environments, where cross-platform portability is essential, this uncertainty compels a fallback to older, less efficient strategies like \emph{Reduce-then-Scan}, sacrificing performance for broader compatibility.

@@

\begin{figure*}[t]
  \centering
  \includegraphics[width=.95\textwidth]{graphics/DecoupledFallback.pdf}
  \caption{\emph{Decoupled Fallback}. For brevity, the chained propagation phase is condensed, see \ref{fig:decoupled-lookback} for the full behavior.}
  \label{fig:decoupled-fallback}
\end{figure*}

@@

@@

We now detail the \emph{Decoupled Fallback} procedure, presenting the algorithm from the perspective of a single workgroup. The details of tile state encoding and atomic updates are discussed in the following section and may vary depending on the data type, so for simplicity, we describe the operation as being performed by the \emph{first thread} of the workgroup:
\begin{enumerate}
  \item[(0)] \textbf{Posting Local Reduction:}
        \begin{itemize}
          \item Each workgroup posts its local reduction---the result of the workgroup-wide scan over partials---to global memory and updates its tile state to \emph{Ready}.
          \item If it is the first work tile, its local reduction \emph{is} its inclusive reduction. In this case, the workgroup updates its tile state directly to \emph{Inclusive} and bypasses the \emph{Chained Propagation} phase entirely.
        \end{itemize}

@@

\begin{figure}
  \centering
  \includegraphics[width=\columnwidth]{graphics/FlowChart.pdf}
  \caption{Flow Chart Representation of \emph{Decoupled Fallback}.\label{fig:decoupled_fallback}}
\end{figure}
